---
format: docx
execute: 
  echo: false
  warning: false
  error: false
---

```{r}
#| label: Package Install and Load
#| output: false

## Install and load packages

# Create vector of packages

requiredPackages <- c("tidyverse", "readxl", "gt", "sf", "ggrepel", "knitr")

# Function to install and load any packages not installed

ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
  install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

# Load packages

ipak(requiredPackages)
```

```{r}
#| label: Data Import
#| output: false

# Read excel data from the Projects "Data" folder. 

data <- read_excel("../../Data/Western_Snowy_Plover_Window_Survey_dataset.xlsx") # Read in our downloaded plover dataset.
head(data) # Show the first few rows of data.

# Read shapefile data from the Projects "Data" folder. 

boundary_sf <- st_read(dsn = "../../Data/us-state-boundaries")
```

## U.S. Fish and Wildlife Service

### Western Snowy Plover Exercise - Master

#### Annual Report

##### Survey Summary

A total of `r nrow(data)` individual western snowy plover (**Charadrius nivosus nivosus**) surveys were conducted during the annual U.S. Fish and Wildlife Service seasonal west coast plover survey. Surveys were conducted at `r combine_words(unique(data$Site))` sites, historically surveyed locations during the survey. In total, `r sum(data$Total)` plovers were observed during the survey over the course of `r round(sum(data$'Duration (Hours)'), 2)` observational hours.

##### Introduction

The western snowy plover (**Charadrius nivosus nivosus**), a small shorebird native to the Pacific coast, has been designated as a threatened species under the Endangered Species Act due to habitat loss, predation, and human disturbance. This species primarily inhabits sandy beaches and coastal dunes, where it nests in the intertidal zone and feeds on invertebrates. Characterized by its cryptic plumage, the western snowy plover is particularly vulnerable during the breeding season, which typically occurs from March to September. Observations conducted by the U.S. Fish and Wildlife Service along select beaches in the Pacific Northwest have provided critical data on population dynamics, nesting success, and habitat utilization, contributing to the understanding of the species' ecological needs and challenges.

Annual observation surveys are essential for informing management strategies aimed at conserving the western snowy plover and its habitat. These surveys not only track population trends and breeding success but also assess the impact of environmental changes and human activities on their nesting sites. By analyzing the collected data, wildlife managers can implement targeted conservation measures, such as habitat protection, predator control, and public education initiatives, to mitigate threats to this endangered species. Continued monitoring and research are vital to ensuring the long-term survival of the western snowy plover along the Pacific Northwest coast.

##### Methods

The annual surveys of western snowy plovers employed systematic observational methods to ensure accurate counts and assessments of the population. Observers conducted visual counts using binoculars while walking along designated survey areas, which included sandy beaches and the lower portions of rivers and tributaries located in close proximity to the coast (figure 1). This approach allowed for a thorough examination of the habitat where snowy plovers are typically found, enhancing the likelihood of detecting individuals and nests. To maintain consistency and accuracy, survey routes were predetermined, and observers followed established protocols to minimize disturbance to the birds while still capturing essential data.

Observations were conducted during both summer and winter seasons to gather comprehensive information on the breeding and migratory populations of western snowy plovers at the survey sites. During the summer months, surveys focused on identifying nesting sites and assessing breeding success, while winter surveys aimed to document the abundance and distribution of migratory plovers that utilize these coastal environments. 

```{r}
#| label: Survey Site Map
#| fig-cap: "Figure 1. Map of Western Snowy Plover survey locations along the coastline."
#| output: asis

Plover_points <- data %>%
  group_by(Site) %>% # Groups records by Area
  summarise(Avg_Lat = mean(y, na.rm = TRUE), Avg_Lon = mean(x, na.rm = TRUE)) # Creates averages of geometry for each Area

### Converts DF into sf object
# Summary: This line of code takes the dataframe that was created above and turns it into a sf object that can be used for plotting on the map.

Plover_sf <- st_as_sf(Plover_points, coords = c("Avg_Lon", "Avg_Lat"), crs = 4326, agr = "constant") # Creates a sf object of the above DF

### Plots Sampling Map

Site_map <- ggplot() +
  geom_sf(data = boundary_sf, size = .5, color = "black",fill="grey90") + # Plots the Great Lakes Polygon shapefile
  geom_sf(data = Plover_sf, size = 3, shape = 21, fill = "red3", color = "red1", stroke = 1.25, alpha = .5) +
  geom_sf_text(data = boundary_sf, aes(label = name)) +
  geom_text_repel(data = Plover_points, aes(x = Avg_Lon, y = Avg_Lat, label = Site), size = 3.1, point.padding = .75, max.overlaps = Inf) + 
  coord_sf(xlim = c(-128.5, -117.2), ylim = c(46.85, 38.1), expand = FALSE) + # Zooms in on the specified area of the Great Lakes shapefile
  #annotation_scale(location = "tr", style = "ticks", line_col = "black", text_col = "black") + # Applies a scale bar to the plot
  theme(panel.background = element_rect(fill = "#c9e9fa"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black"), axis.title.y = element_blank(), axis.title.x = element_blank(), axis.line = element_line(colour = "black"), panel.border = element_rect(colour = "black", fill=NA, size=.5)) # Gives the plot a plain standard appearance

Site_map # calls the map to be plotted
```

##### Results

The data on Western Snowy Plover total counts indicate notable seasonal and site-specific trends. In the summer, Leadbetter Point had the highest population, with a total count of 85 individuals, followed by Agate Beach with 56, while Bayshore Spit had the lowest at 24 (table 1). This suggests that Leadbetter Point and Agate Beach are critical breeding areas. In contrast, winter counts showed a significant decline across all sites, with Agate Beach recording 25 plovers, Leadbetter Point 21, and Bayshore Spit 19. The decrease in overall counts during the winter season, especially the absence of juveniles, highlights the migratory behavior of the plovers and underscores the importance of these habitats for both breeding and wintering populations.

Furthermore, observational plover counts per hour reveal significant seasonal trends across the sites. In the summer, Leadbetter Point had the highest observation rate at 10.97 plovers per hour, indicating a robust presence during the breeding season, followed closely by Agate Beach with 8.3 observations per hour (figure 2). Bayshore Spit had the lowest summer observation rate at 3.0, suggesting it is less frequented by plovers during this time. Conversely, winter counts showed a marked reduction in observational rates across all sites, with Agate Beach recording 3.33 observations per hour, Leadbetter Point at 2.9, and Bayshore Spit at 2.05. This decline underscores the migratory patterns and decreased visibility of the plovers during the non-breeding season, highlighting the importance of seasonal monitoring to understand the dynamics of plover populations in these habitats.

```{r}
#| label: Plover Observation Summary Table
#| fig-cap: "Table 1. Seasonal Western Snowy Plover observational counts by survey site."

# Group by Survey Season and Site, and summarize Total Plover counts
Obs_data <- data %>%
  group_by(`Survey Season`, Site) %>%
  summarise(Total_Males = sum(Males),
            Total_Females = sum(Females),
            Total_Juveniles = sum(Juveniles),
            Total_Count = sum(Total),
            .groups = 'drop') %>%
  arrange(Site)  # Arrange by Site

# Create a gt table
Obs_table <- Obs_data %>%
  gt() %>%
  cols_label(
    Total_Males = "Total Males",
    Total_Females = "Total Females",
    Total_Juveniles = "Total Juveniles",
    Total_Count = "Total Count"
  )

# Print the table
Obs_table
```

```{r}
#| label: Plover Observation per Hour Plot
#| fig-cap: "Figure 2. Text."

# Group by Survey Season and Site, and summarize Total Plover counts by duration (hours)
Effort_data <- data %>%
  group_by(`Survey Season`, Site) %>%
  summarise(Obs_Hour = round(sum(Total)/sum(`Duration (Hours)`), 2)) %>%
  arrange(Site)  # Arrange by Site

# Create a ggplot
Effort_plot <- Effort_data %>%
  ggplot(., aes(x = Site, y = Obs_Hour, fill = `Survey Season`)) +
  geom_bar(stat = "identity", color = "black", position = position_dodge()) +
  scale_fill_manual(values = c('#e5a300','#0487ee')) +
  labs(x = "", y = "Plover Observations per Hour") +
  theme_classic() +
  theme(axis.text = element_text(color = "black", size = 10))

# Print the plot
Effort_plot
```

##### Discussion

The results from the observational and total count data of Western Snowy Plovers indicate significant seasonal and site-specific trends that are crucial for understanding the species' population dynamics. The higher summer counts and observational rates at Leadbetter Point and Agate Beach suggest these locations are vital breeding habitats, supporting robust populations during the nesting season. The substantial presence of juveniles at these sites further emphasizes their importance for reproductive success. Conversely, the notable decline in both total counts and hourly observations during the winter months across all sites highlights the migratory nature of the plovers and their reduced visibility during this period. Such seasonal variations in abundance and behavior reinforce the need for targeted conservation efforts during critical breeding times, as well as continued monitoring throughout the year to address the challenges posed by habitat loss and human disturbances. These findings underscore the importance of maintaining and enhancing habitat quality at key sites to support the long-term survival of the western snowy plover in the Pacific Northwest.